---
title: "MorteCovid"
date: "`r Sys.Date()`"
---

## Covid death among hospitalized children and young in 2022

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, error = FALSE, warning = FALSE, message = FALSE)

#Carregar bibliotecas necessárias
library(readr)
library(dplyr)
library(tidytable)
#library(table1)
library(tableone)
library(ggplot2)
library(pyramid)
library(lubridate)
library(kableExtra)
library(finalfit)
library(expss)
library(geobr)
library(sf)

#Carregar arquivo com a base de dados
#substituir pelo arquivo de vocês INFLUD22-31-10-2022.csv

SIVEP <- read_delim("INFLUD22-17-10-2022.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE)
names(SIVEP) <- tolower(names(SIVEP))
```

## Time distribution of Covid deaths by epi week in 2022
```{r echo=FALSE}
TempSIVEP <- SIVEP %>% 
  filter.(classi_fin == 5) %>% 
  mutate.(obito = case_when.(evolucao == 2 ~ 1,
                             evolucao == 1 ~ 0)) %>% 
  select.(obito, dt_evoluca) %>% 
  filter.(obito == 1) %>% 
  mutate.(dt_evoluca = dmy(dt_evoluca)) %>% 
  mutate.(semana = week(dt_evoluca)) %>% 
  summarize.(Obitos = sum(obito == 1), .by = semana) 
  ggplot(TempSIVEP) +
  geom_line(aes(x=semana, y = Obitos), size = 2, colour = "#4D77BB") +
  theme(legend.position = c(0.95, 0.15),
        legend.justification = c("right", "top")) +
  theme_classic() +
  labs(
    title = "Deaths by week",
    subtitle = "(2022)",
    caption = "Source: Sivep-Gripe",
    x = "Time (epidemiologial week)",
    y = "Death cases (N)")
```

## Pirâmide etária dos óbitos
```{r echo=FALSE}
ConfSIVEP <- SIVEP %>% 
  filter.(classi_fin == 5) %>% 
  mutate.(obito = case_when.(evolucao == 2 ~ 1,
                             evolucao == 1 ~ 0)) %>% 
  select.(cs_sexo, obito, nu_idade_n, tp_idade, id_municip, dt_interna) %>% 
  filter.(obito == 1) %>% 
  filter.(cs_sexo != "I") %>% 
  filter.(nu_idade_n >= 0) %>% 
  mutate.(idade = case_when(
    tp_idade == 3 & (nu_idade_n >= 0 & nu_idade_n < 5) ~ 1,
    tp_idade == 3 & (nu_idade_n >= 5 & nu_idade_n < 10) ~ 2,
    tp_idade == 3 & (nu_idade_n >= 10 & nu_idade_n < 20) ~ 3,
    tp_idade == 3 & (nu_idade_n >= 20 & nu_idade_n < 30) ~ 4,
    tp_idade == 3 & (nu_idade_n >= 30 & nu_idade_n < 40) ~ 5,
    tp_idade == 3 & (nu_idade_n >= 40 & nu_idade_n < 50) ~ 6,
    tp_idade == 3 & (nu_idade_n >= 50 & nu_idade_n < 60) ~ 7,
    tp_idade == 3 & (nu_idade_n >= 60 & nu_idade_n < 70) ~ 8,
    tp_idade == 3 & (nu_idade_n >= 70 & nu_idade_n < 80) ~ 9,
    tp_idade == 3 & (nu_idade_n >= 80) ~ 10,
    (tp_idade == 1 | tp_idade == 2) & !is.na(nu_idade_n) ~ 1)
  ) %>% 
  mutate.(idade = factor(
    idade, levels = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10), labels = c("1-4y", "5-9y", "10-19y", 
                                              "20-29y", "30-39y",
                                              "40-49y", "50-59y",
                                              "60-69y", "70-79y",
                                              "80+"))) %>% 
  mutate.(cs_sexo = factor(cs_sexo, levels = c("F", "M"), labels = c("Mulher", "Homem"))) %>% 
  summarize.(Obitos = sum(obito == 1), .by = c(cs_sexo, idade)) %>% 
  pivot_wider.(names_from = cs_sexo, values_from = Obitos) %>% 
  mutate.(popt = Mulher + Homem) %>% 
  mutate.(PopPerc = popt / sum(popt) * 100) %>% 
  mutate.(PopPercM = Homem / sum(popt) * 100) %>%
  mutate.(PopPercF = Mulher / sum(popt) * 100) %>% 
  select.(PopPercM, PopPercF, idade)
  pyramid(ConfSIVEP, Clab = "Age classes", Llab = "Men (%)", Rlab = "Women (%)", 
        Lcol = "#4d77bb", Rcol = "#ff6555",
        main = "Sex and age structure of Covid deaths in 2022") 
```

## Children and young death proportion distribution in Brazil 2022
```{r echo=FALSE}
MapaBR = read_state(year = 2010, code_state = "all", showProgress = FALSE)
MapMorte <- SIVEP %>% 
  filter.(classi_fin == 5) %>% 
  mutate.(obito = case_when.(evolucao == 2 ~ 1,
                             evolucao == 1 ~ 0)) %>%
  filter.(!is.na(obito)) %>% 
  
      filter.(nu_idade_n >= 0) %>% 
  mutate.(idade = case_when(
    tp_idade == 3 & (nu_idade_n >= 0 & nu_idade_n < 5) ~ 1,
    tp_idade == 3 & (nu_idade_n >= 5 & nu_idade_n < 10) ~ 2,
    tp_idade == 3 & (nu_idade_n >= 10 & nu_idade_n < 20) ~ 3,
    tp_idade == 3 & (nu_idade_n >= 20 & nu_idade_n < 30) ~ 4,
    tp_idade == 3 & (nu_idade_n >= 30 & nu_idade_n < 40) ~ 5,
    tp_idade == 3 & (nu_idade_n >= 40 & nu_idade_n < 50) ~ 6,
    tp_idade == 3 & (nu_idade_n >= 50 & nu_idade_n < 60) ~ 7,
    tp_idade == 3 & (nu_idade_n >= 60 & nu_idade_n < 70) ~ 8,
    tp_idade == 3 & (nu_idade_n >= 70 & nu_idade_n < 80) ~ 9,
    tp_idade == 3 & (nu_idade_n >= 80) ~ 10,
    (tp_idade == 1 | tp_idade == 2) & !is.na(nu_idade_n) ~ 1)
  ) %>% 
  mutate.(idade = factor(
    idade, levels = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10), labels = c("1 yr to 4 yrs", "5 yrs to 9 yrs", "10 yrs to 19 yrs", 
                                              "20 yrs to 29 yrs", "30 yrs to 39 yrs",
                                              "40 yrs to 49 yrs", "50 yrs to 59 yrs",
                                              "60 yrs to 69 yrs", "70 yrs to 79 yrs",
                                              "80 yrs or more"))) %>% 
  filter.(as.integer(idade) < 4) %>% 
  mutate.(total = 1) %>% 
  summarise.(obito = sum(obito == 1), pop = sum(total), .by = ("sg_uf")) %>% 
  filter.(!is.na(sg_uf)) %>% 
  mutate.(propobito = round((obito/pop)*100, 1)) %>% 
  mutate.(code_state = case_when.(
    sg_uf == "RO" ~ 11, sg_uf == "AC" ~ 12, sg_uf == "AM" ~ 13, sg_uf == "RR" ~ 14, sg_uf == "PA" ~ 15,
    sg_uf == "AP" ~ 16, sg_uf == "TO" ~ 17, sg_uf == "MA" ~ 21, sg_uf == "PI" ~ 22, sg_uf == "CE" ~ 23,
    sg_uf == "RN" ~ 24, sg_uf == "PB" ~ 25, sg_uf == "PE" ~ 26, sg_uf == "AL" ~ 27, sg_uf == "SE" ~ 28,
    sg_uf == "BA" ~ 29, sg_uf == "MG" ~ 31, sg_uf == "ES" ~ 32, sg_uf == "RJ" ~ 33, sg_uf == "SP" ~ 35,
    sg_uf == "PR" ~ 41, sg_uf == "SC" ~ 42, sg_uf == "RS" ~ 43, sg_uf == "MS" ~ 50, sg_uf == "MT" ~ 51,
    sg_uf == "GO" ~ 52, sg_uf == "DF" ~ 53,)) %>% 

  inner_join.(MapaBR, .by = "code_state") %>% 
  st_as_sf()
  
  ggplot(data = MapMorte) +
  geom_sf(aes(fill = propobito ), size = .15) +
  scale_fill_gradient(low = "light blue", high = "dark blue", name = "Death proportion") +
  #geom_sf_text(aes(label = name_state), colour = "white") +
  xlab("") +  ylab("") +
  theme_classic()

```

## Risk factors for Covid death among children and young in Brazil, 2022
```{r echo=FALSE}
  DescSIVEPult <- SIVEP %>% 
  filter.(classi_fin == 5) %>% 
  filter.(cs_sexo != "I") %>% 
  #filter.(month(dt_sin_pri) >= 3) %>% 
  mutate.(obito = case_when.(evolucao == 2 ~ 1,
                             evolucao == 1 ~ 0)) %>% 
  filter.(nu_idade_n >= 0) %>% 
  mutate.(idade = case_when(
    tp_idade == 3 & (nu_idade_n >= 0 & nu_idade_n < 5) ~ 1,
    tp_idade == 3 & (nu_idade_n >= 5 & nu_idade_n < 10) ~ 2,
    tp_idade == 3 & (nu_idade_n >= 10 & nu_idade_n < 20) ~ 3,
    tp_idade == 3 & (nu_idade_n >= 20 & nu_idade_n < 30) ~ 4,
    tp_idade == 3 & (nu_idade_n >= 30 & nu_idade_n < 40) ~ 5,
    tp_idade == 3 & (nu_idade_n >= 40 & nu_idade_n < 50) ~ 6,
    tp_idade == 3 & (nu_idade_n >= 50 & nu_idade_n < 60) ~ 7,
    tp_idade == 3 & (nu_idade_n >= 60 & nu_idade_n < 70) ~ 8,
    tp_idade == 3 & (nu_idade_n >= 70 & nu_idade_n < 80) ~ 9,
    tp_idade == 3 & (nu_idade_n >= 80) ~ 10,
    (tp_idade == 1 | tp_idade == 2) & !is.na(nu_idade_n) ~ 1)
  ) %>% 
  mutate.(idade = factor(
    idade, levels = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10), labels = c("1 yr to 4 yrs", "5 yrs to 9 yrs", "10 yrs to 19 yrs", 
                                              "20 yrs to 29 yrs", "30 yrs to 39 yrs",
                                              "40 yrs to 49 yrs", "50 yrs to 59 yrs",
                                              "60 yrs to 69 yrs", "70 yrs to 79 yrs",
                                              "80 yrs or more"))) %>% 
  filter.(as.integer(idade) < 4) %>% 
  mutate.(cs_sexo = factor(cs_sexo, levels = c("F", "M"), labels = c("Women", "Men"))) %>% 
  select.(obito, cs_sexo, idade, cs_raca, cs_escol_n, cardiopati,
          hematologi, sind_down, hepatica, asma, diabetes, neurologic, pneumopati, imunodepre,
          renal, obesidade, dt_1_dose, dt_2_dose, uti, dt_entuti, dt_saiduti, suport_ven, 
          vacina_cov, dose_1_cov, dose_2_cov, dose_ref, fab_cov_1, fab_cov_2, fab_covref, 
          sg_uf) %>% 
  mutate.(cs_raca = factor(cs_raca, levels = c(1, 2, 3, 4, 5), 
                           labels = c("White", "Black", "Asian", "multiracial", "Indigenous"))) %>% 
  mutate.(cs_escol_n = factor(cs_escol_n, levels = c(0, 1, 2, 3, 4, 5), 
                           labels = c("Illiterate", "Basic 1st cycle", 
                                      "Basic 2nd cycle", "High scholl", "College", 
                                      "Not applicable"))) %>% 
  mutate.(cardiopati = factor(cardiopati, levels = c(2, 1), 
                           labels = c("No", "Yes"))) %>%
  mutate.(hematologi = factor(hematologi, levels = c(2, 1), 
                           labels = c("No", "Yes"))) %>%
  mutate.(sind_down = factor(sind_down, levels = c(2, 1), 
                           labels = c("No", "Yes"))) %>%
  mutate.(hepatica = factor(hepatica, levels = c(2, 1), 
                           labels = c("No", "Yes"))) %>%
  mutate.(asma = factor(asma, levels = c(2, 1), 
                           labels = c("No", "Yes"))) %>%
  mutate.(diabetes = factor(diabetes, levels = c(2, 1), 
                           labels = c("No", "Yes"))) %>%
  mutate.(neurologic = factor(neurologic, levels = c(2, 1), 
                           labels = c("No", "Yes"))) %>%
  mutate.(pneumopati = factor(pneumopati, levels = c(2, 1), 
                           labels = c("No", "Yes"))) %>%
  mutate.(imunodepre = factor(imunodepre, levels = c(2, 1), 
                           labels = c("No", "Yes"))) %>%
  mutate.(renal = factor(renal, levels = c(2, 1), 
                           labels = c("No", "Yes"))) %>%
  mutate.(obesidade = factor(obesidade, levels = c(2, 1), 
                           labels = c("No", "Yes"))) %>%
  mutate.(vacina_cov = factor(vacina_cov, levels = c(2, 1), 
                           labels = c("No", "Yes"))) %>%
  mutate.(uti = factor(uti, levels = c(2, 1), 
                           labels = c("No", "Yes"))) %>% 
  mutate.(vac1 = case_when.(!is.na(dose_1_cov) ~ 1,
                            is.na(dose_1_cov) ~ 0)) %>% 
  mutate.(vac2 = case_when.(!is.na(dose_2_cov) ~ 1,
                            is.na(dose_2_cov) ~ 0)) %>% 
  mutate.(vacref = case_when.(!is.na(dose_ref) ~ 1,
                            is.na(dose_ref) ~ 0)) %>% 
  
  mutate.(vacref = factor(vacref, levels = c(0, 1), labels = c("No", "Yes"))) %>% 
  mutate.(dt_entuti = dmy(dt_entuti)) %>% 
  mutate.(dt_saiduti = dmy(dt_saiduti)) %>% 
  mutate.(tp_uti = as.numeric(dt_saiduti - dt_entuti)) %>% 
  mutate.(suport_ven = factor(suport_ven, levels = c(1, 2, 3), 
                              labels = c("Invasive", "Non invasive", "No vent. support"))) %>% 
  mutate.(obito = factor(obito, levels = c(0, 1), labels = c("No", "Yes"))) %>% 
  rename.(Death = obito) %>% 
  rename.(Sex = cs_sexo) %>% 
  rename.(Age = idade) %>% 
  rename.(Race = cs_raca) %>% 
  rename.(Scholarship = cs_escol_n) %>% 
  rename.(`Heart disease` = cardiopati) %>% 
  rename.(`Hematological disease` = hematologi) %>% 
  rename.(`Down syndrome` = sind_down) %>% 
  rename.(`Liver disease` = hepatica) %>% 
  rename.(Ashtma = asma) %>% 
  rename.(Diabetes = diabetes) %>% 
  rename.(`Neurological disase` = neurologic) %>% 
  rename.(`Lung disease` = pneumopati) %>% 
  rename.( Immunosupression = imunodepre) %>% 
  rename.(`Kidney disease` = renal) %>% 
  rename.(Obesity = obesidade) %>% 
  rename.(UCI = uti) %>% 
  rename.(`Time at UCI` = tp_uti) %>%
  rename.(`Ventilatory support` = suport_ven) %>%
  rename.(`Covid vaccine` = vacina_cov) %>%
  rename.(`1st dose` = vac1) %>%
  rename.(`2nd dose` = vac2) %>%
  rename.(`Booster dose` = vacref)
  
BaseVar <- c("Sex", "Race", "Age", "Scholarship", "Heart disease", "Hematological disease", "Down syndrome", 
             "Liver disease", "Ashtma", "Diabetes", "Neurological disase", "Lung disease", "Immunosupression", 
             "Kidney disease", "Obesity", "UCI", "Time at UCI", "Ventilatory support", "Covid vaccine", "Booster dose")
BaseVarCat <- c("Sex", "Race", "Age", "Scholarship", "Heart disease", 
  "Hematological disease", "Down syndrome", "Liver disease", "Ashtma", "Diabetes", "Neurological disase", "Lung disease",
  "Immunosupression","Kidney disease", "Obesity", "UCI", "Ventilatory support", "Covid vaccine", "Booster dose")
Tab2 <- CreateTableOne(vars = BaseVar, strata = "Death" , data = DescSIVEPult, factorVars = BaseVarCat)
#print(Tab1, nonnormal = "tp_uti", formatOptions = list(big.mark = ","))
kableone(Tab2)

```

## Multiv. analysis for death in children - OR Plot
```{r children, echo=FALSE}
explaninf <- c("Sex", "UCI", "Obesity", "Immunosupression") 
desf1inf <- "Death"
# Tabdesf <- DescSIVEPult %>% 
# finalfit(desf1inf, explaninf) 
# kable(Tabdesf)

DescSIVEPult %>%
  or_plot(desf1inf, explaninf, remove_ref = T, table_text_size=3, title_text_size=12,
          plot_opts=list(xlab("OR, 95% CI"), 
                         theme(axis.title = element_text(size=9))))

```

## <span style="color: red;">Desafio: refazer a tabela 1 e o forest plot de um modelo para mulheres incluindo idade como fator de risco</span>
```{r women, echo=FALSE}

```
